---
title: "Weather Data Analysis"
author: "Daniel Starer"
date: "2025-07-24"
output:
  word_document: default
  html_document: default
  pdf_document: default
upd: "2025-07-25"
---

A bit of context first.  This data originates from Zaruhi Avagyan on Kaggle
and no information was provided by the creator regarding what locale
it originates from.  The data, nonetheless, is very detailed and I wanted to
visualize the correlation between humidity and precipitation and temperature and precipitation.

First, we will load the dataset.
```{r}
my_weather <- read.csv("weather.csv")
```

Next, to best show correlation, we will need to use ggplot2, so we'll load that library.
```{r}
library(ggplot2)
```

Now, let's take a look at the data to get a sense of the data's structure.
```{r}
head(my_weather)
```

There are multiple columns for humidity and temperature because there are two times
in which humidity and temperature were recorded: 9 AM and 3 PM.  Because of this, we
will need to do some cleaning to organize things a bit.  We can achieve this by
combining the data in both humidity columns and the data in both temperature
columns and averaging them out to find the average temperature and humidity.

```{r}
my_weather$AvgHumidity <- rowMeans(my_weather[, c("Humidity9am", "Humidity3pm")], na.rm = TRUE)
my_weather$AvgTemp <- rowMeans(my_weather[, c("Temp9am", "Temp3pm")], na.rm = TRUE)
AvgHumidity <- my_weather$AvgHumidity
AvgTemp <- my_weather$AvgTemp
```

Now that we have separated the average humidity and average temperature into their
own objects, we can create an object for rainfall and test the correlation between
these objects with linear models.
```{r}
Rainfall <- my_weather$Rainfall
```

```{r}
plot(AvgHumidity, Rainfall, pch = 16, col = "blue", main = "Rainfall plotted against Average Humidity", xlab = "Avg. Humidity", ylab = "Rainfall")
lm(Rainfall ~ AvgHumidity)
abline(lm(Rainfall ~ AvgHumidity))
```

```{r}
plot(AvgTemp, Rainfall, pch = 16, col = "blue", main = "Rainfall plotted against Average Temperature", xlab = "Avg. Temp.", ylab = "Rainfall")
lm(Rainfall ~ AvgTemp)
abline(lm(Rainfall ~ AvgTemp))
```

From these visualizations, we can see there is a general correlation between humidity and rainfall
and temperature and rainfall, however there are a fair amount of outliers.  Instances where
there is no recorded rainfall especially conforms more to the humidity and temperature.

